## Section III: Hardware Implementation

### 3.1 Evidence boundary

The RTL is portable Verilog and expresses the $A_{31}$ multiplier as sixteen
logical 32-by-32 products followed by 72-bit accumulation and Mersenne
reduction. It does not instantiate DSP48E1 cascade, pattern-detect, or carry
cascade primitives. Physical DSP usage is therefore a property of a named
synthesis image, not the source-level product count.

| Artifact | Logic / registers | DSP / BRAM | Timing and status |
|:---------|:------------------|:-----------|:------------------|
| Wukong Artix-7 100T `RPLU2PADE` | 20,277 / 126,800 `SLICE_LUTX`; 6,678 / 126,800 `SLICE_FFX` | 72 / 240 DSP48E1; 0 BRAM | `clk_fast` max 36.54 MHz; routed and silicon-tested at 2 MHz |
| Tang GW5A-25A `rplu2_arith_probe` | 9,211 LUT4; 7,926 DFF; 822 ALU | 0 DSP; 0 BRAM | reported clocks max 54.42/48.40 MHz; routed and silicon-tested at 12 MHz |

The Tang artifact covers RPLU2 configuration plus QLDI/QSUB arithmetic. It is
not a full Padé evaluator and must not be used as a full-pipeline utilization
number. The authoritative run details are in `docs/hardware_evidence.md`.

### 3.2 Shared multiplier organization

For jet order $n=2$, the Cauchy product contains six $A_{31}$ multiplies. A
single shared multiplier time-division-multiplexes those operations. Each
$A_{31}$ multiply contains sixteen logical integer products, so a hypothetical
six-way unrolled model contains 96 logical products. Neither number is a count
of physical DSP48E1 cells. The routed Wukong image uses 72 DSP48E1 cells for
the complete `RPLU2PADE` artifact.

### 3.3 M31 reduction

The implemented reduction is target-independent RTL. It splits each 72-bit
accumulator into 31-bit chunks, adds the chunks, and conditionally subtracts
$p=2^{31}-1$. Claims about DSP48E1 pattern-detect or zero-overhead cascade
routing are intentionally excluded because the current RTL does not implement
those primitives.

### 3.4 Integration scope

The Artix-7 100T is the constrained-integration and silicon-evidence target for
the live Padé evaluator. The Tang 25K remains the subsystem probe platform.
Full concurrent builds containing RPLU2 plus additional sidecars and safety
layers are an Artix-7 200T / Kintex-class direction rather than a demonstrated
100T or Tang resource result.
